Whenever solving a problem like this, it's always good to build it up one step at a time. You could be tempted to jump right to the solution but it is easier to take small steps and check your results along the way than *debugging* a long piece of code.



#### Divisible by 4

The first step is to implement the simplest of the logical checks: is the year divisible by 4? If it is, print out an appropriate message checking that it works for a know leap year.

```{r}
#| filename: leap_year.r
year <- 2024

if (year %% 4 == 0) {
    cat(year, "is a leap year")
} 
```

For a year that we know is not a leap year, what do we get:

```{r}
#| filename: leap_year.r
year <- 2023

if (year %% 4 == 0) {
    cat(year, "is a leap year")
}
```


We get no output, so let's fix that next, by adding an `else`:

```{r}
#| filename: leap_year.r
year <- 2024

if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```

```{r}
#| filename: leap_year.r
year <- 2023

if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```





So far it's looking like it's doing the right thing. We have a working leap year calculator which covers most cases correctly.

#### Divisible by 100

Let's check the next corner case: what happens if the year is divisible by 100. The rules say that this should not be a leap year:

```{r}
#| filename: leap_year.r
year <- 1900

if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```


So it's currently getting this wrong. We need to add in a check to see if the year is divisible by 100 and set it to be labelled as not a leap year.

Here we need to be careful as any number which is divisible by 100 is also divisible by 4 but we want a special thing to happen in the former case. The divisible-by-100 check is more specific than the divisible-by$4$ check, so it must come first in the `if`-`else if`-`else` chain:

```{r}
#| filename: leap_year.r
year <- 2024

if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```

*(omitting repeated code)*
```{r}
#| filename: leap_year.r
#| echo: false
year <- 2023

if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```


```{r}
#| filename: leap_year.r
#| echo: false
year <- 1900

if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```


#### Divisible by 400

The code is working for all those cases correctly, however, the last condition to think about is what happens for the year 2000:


```{r}
#| filename: leap_year.r
year <- 2000

if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```


This year should be a leap year as it is divisible by 400. Again this is more specific than the previous two so must be added before them in the chain:

```{r}
#| filename: leap_year.r
year <- 2024

if (year %% 400 == 0) {
    cat(year, "is a leap year")
} else if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```


*(omitting repeated code)*
```{r}
#| filename: leap_year.r
#| echo: false
year <- 2023

if (year %% 400 == 0) {
    cat(year, "is a leap year")
} else if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```

```{r}
#| filename: leap_year.r
#| echo: false
year <- 1900

if (year %% 400 == 0) {
    cat(year, "is a leap year")
} else if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```

```{r}
#| filename: leap_year.r
#| echo: false
year <- 2000

if (year %% 400 == 0) {
    cat(year, "is a leap year")
} else if (year %% 100 == 0) {
    cat(year, "is not a leap year")
} else if (year %% 4 == 0) {
    cat(year, "is a leap year")
} else {
    cat(year, "is not a leap year")
}
```


