---
title: Hello, R!
---

R is a scripting language, meaning that commands are interpreted at runtime rather than translated into machine code (compiled) beforehand. The simplest way to run a command in R is making use of the R command line. Follow these steps:

- Open a Command Prompt (Windows) or Terminal (MacOS or Linux) on your computer. Alternatively, you can try this on the tab *Terminal* in RStudio.
- Type `R` and press *Enter* to start the R interpreter. You should see the R prompt (`>`).

```{.bash filename="Terminal/Command Prompt"} 
R
```
```{r filename="Terminal/Command Prompt"}
#| echo: false
cat("R version 4.4.0 (2024-04-24) -- \"Puppy Cup\"
Copyright (C) 2024 The R Foundation for Statistical Computing
Platform: aarch64-apple-darwin20

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

> ") 
```

- At the prompt, type a simple R command like `print("Hello, R!")` or `cat("Hello, R!")`. Press *Enter* to execute the command. You should see "Hello, R!" printed on the screen.
```{.r}
cat("Hello, R!")
```

**You can do the same directly on the tab *Console* in RStudio**. In fact, RStudio just did the step of opening R for us.

To exit the R interpreter, type `q()`. Note, that if you do this in RStudio you will close the program.



## Scripts


A R script is a simple text file containing a set of instructions. Scripts allow us to save instructions to automate repetitive tasks, streamline processes, and perform specific functions without manual intervention. This is called batch processing. R will read your script, starting at the top and running each line of code until it reaches the bottom.

You can write script files using any text editor. Here we are going to use **RStudio**, so let's start by going to the text editor panel in RStudio and type into it:

```{.r}
cat("Hello, R!")
```

Then go to *File → Save* and give it a more descriptive name. R files end in the extension ".R" so change its name to "script.R". In the far right panel you will see a list of files, and you should be able to see `script.R`.

To run the script that we've just written and saved, we move to the editor and click the *Run* button on the top right side. This will execute the command in the current line or selected lines. You can also write or paste commands directly on the R Console tab, try it as well.

::: {#keypoints-N .callout-tip icon=false title='Note that'}
- Always give your R scripts a ".R" extension
- In RStudio, R *code* goes in the text editor window while you *run* the code on the console or the terminal tabs.
:::

Alternative, we can use the Terminal to run a script. The only command we need to know about here is `RScript` which we run by typing `Rscript` in that window, followed by a space, followed by the name of our script file (including the ".R" extension) and then pressing *Enter*:

```{.bash filename="Terminal/Command Prompt"} 
RScript script.R
```

If you do that you should see, printed to the terminal:

```{r}
#| filename: "script.R" 
cat("Hello, R!")
```

Congratulations, you've just written and executed your very first R script!


## Commenting your script

It is **good practice** to add comments to your code so it is easy to understand by other programmers or even yourself after a few months. This is particularly important if there's anything non-obvious in your code. To add a comments you can use a `#` symbol at any position in your script, what is behind '#' will be ignored by R when running that line of code. 

```{r filename="colour.r"}
# This script asks for your favourite colour
fav <- "blue"

# cat("Don't print me!") 

cat("My favourite colour is", fav) # print my favourite colour
```





