---
title: Basic data types
---

In R we have several built-in basic data types. The most common types of data that you will find are numbers, characters and booleans.

::: {#keypoints2 .callout-tip icon=false title='Key points'}
- R is dynamically typed, meaning you don't need to declare the type of a variable explicitly.
- You can use the `class()` function to check the data type of a variable.
- Data types determine what operations can be performed on the data.
- You can convert between different data types using built-in functions like `as.integer()`, `as.numeric()` and `as.character()`
:::

### Numeric Types

Description                | Type     | Example 
---------------------------|----------|----------------
integers or whole numbers  | integer  | `a <- as.integer(12)`
floating-point numbers     | numeric  | `b <- as.numeric(7.3)`
complex numbers            | complex  | `c <- as.complex(2i + 7)`


```{.r}
a <- 12L
b <- 7.3
c <- 2i + 7

cat("A:", a, class(a))
cat("B:", b, class(b))
cat("C:", c, class(c))
```

```{r}
#| echo: false
a <- 12L
b <- 7.3
c <- 2i + 7

cat("A:", a, class(a))
cat("B:", b, class(b))
cat("C:", c, class(c))
```


### Text Types

Description                       | Type       | Example 
----------------------------------|------------|----------------
textual data (strings)            | character  | `a <- "Jean Golding"`

```{r}
name <- "Jean Golding"
cat("Name:", name, class(name))
```

::: {#keypoint1 .callout-tip icon=false title="Note that"}
It’s important that when writing numbers in your scripts, you do not put quotation marks around them, otherwise they will be recognized by the R interpreter as strings. There is a difference between 3.14159 and "3.14159", the first is a number and the second is just a string of characters.

```{r}
pi <- 3.14159
cat("pi:", pi, class(pi))
```
```{r}
pi <- "3.14159"
cat("pi:", pi, class(pi))
```
:::


### Boolean Types

Description                       | Type     | Example 
----------------------------------|----------|----------------
boolean values (True or False)    | boolean  | `a <- TRUE`

```{r}
a <- 23
b <- 2
c <- a != b

cat("C:", c, class(c))
```



::: {#exercise-1 .callout-note title="Exercise" icon=false} 

Without using R, can you tell what is the data type of these variables?
```{.r}
x <- 32
number_of_participants <- "1017"
Friday <- TRUE
y <- 1L
a <- 10<8
```
:::

::: {#answer .callout-caution icon=false title='Answer' collapse="true"}
`x` is a numeric (float)

`number_of_participants` is a string

`Friday` is a boolean

`y` is an integer 

`a` is a boolean
:::
