---
title: While loop
---

We have seen that `for` loops can repeatedly execute a block of code for each element specified. We can think of another scenario where we only want to continue executing a block of code as long as a given condition is `TRUE`. The basic idea is:
```
    WHILE there is money in my account
        have lunch out
        update my bank statement
```


In `while` loops the condition is checked at the beginning of each iteration and the loop continues executing as long as the condition is `TRUE`. The basic syntax in R which has a similar scaffolding to an `if` clause:
```{.R}
while (condition) {
    # code to execute while condition is True
}
```


To ensure that the condition will eventually become `FALSE`, it is common usage to increment/decrement a counter variable in the loop body:

```{r}
count <- 0
while ( count < 5 ) {
    cat(count, " ")
    count <- count + 1
}
```

There are some extra flow control instructions in `while` loops that we are not going to see here in detail but it is important to mention them. They are:

- `break`: exits the loop immediately
- `next`: skips the rest of the current iteration and moves to the next

```{.R}
while (TRUE) {
    user_input <- readline("Enter 'q' to quit: ")
    if (user_input == 'q') {
        break
    }
}
```

::: {#exercise1 .callout-note title="Exercise" icon=false} 
Given a list of students **sorted** by name

```{.R}
students <- c("Alice", "Bob", "Charlie", "Frank", "George", "Hannah", "Julia", "Mia", "Noah", "Olivia")
```

create a script `while.R` that prints the names of students starting with letters before 'M' in the alphabet. 
:::

::: {#answer1 .callout-caution icon=false title='Answer' collapse="true"}

Note in the below script that the `while` loop needs to take into account two different conditions:

- That the student's name starts with a letter before 'M'
- That we don't try to index an element out of range in `students`

```{r}
students <- c("Alice", "Bob", "Charlie", "Frank", "George", "Hannah", "Julia", "Mia", "Noah", "Olivia")

# Initialize the index
i <- 1

# While loop to print names before 'M'
while (i<length(students) & students[i]<"M") {
    cat(students[i], " ")
    i <- i + 1
}
```

:::




